A radiocarbon data based evaluation of Cultural Transmission in Bronze Age burial rites of Central, Northern and North-western Europe

### 2019-02-02 article/03_Results.Rmd
more consistent spacing before "calBC" (@yesdavid)

In the course of the Bronze Age in Europe there was a trend away from body burial towards cremation. Around 2200calBC cremation burials were almost unknown in Poland, Southern Germany, Northeastern France and Northern Germany. In the Northwest, in Great Britain and the Benelux area, and in the Southeast, in Austria and the Czech Republic, however, there were early cremation traditions. These contexts could be discussed as areas of origin of the later omnipresent phenomenon. While cremation burials lost importance in Austria and the Czech Republic as well as in the Benelux region during the Early Bronze Age and only regained relevance there in the Middle Bronze Age, their share increased steadily in England. Inhumation and cremation were balanced here for a long time. The situation was similar in Southern Scandinavia, where the proportion of cremation burials was much lower until the Late Bronze Age. In Poland and northern Germany, cremation became relatively suddenly and completely established with the beginning of the Late Bronze Age. In Northeastern France and Northern Germany this change took place already in the Middle Bronze Age. In both contexts inhumation continued to play an important role.

@@

Beyond the simple comparison of archaeological knowledge and information derived from Radon-B, Figure \ref{sed_region_matrix} shows the time series of squared Euclidean distance for each region-region relationship. The majority of the curves in its top right half for *burial type* correspond to expectations: At the beginning of the Bronze Age, inhumation is widespread throughout the study area, so the Euclidean distance at 2200calBC is small in almost every region-region relationship. Only the Benelux countries, where the share of cremation is already high in the Early Bronze Age, are excluded. From the Middle Bronze Age onwards, but especially at the transition to the Late Bronze Age, the transition to the dominance of cremation took place all over Europe. The SED values increase in this period and show clear peaks or several successive maxima. This rise is a consequence of the time lag in the transition processes as depicted in Radon-B: It is different in every region. At the end of the Late Bronze Age, the average SED is approaching zero again, as burial by cremation dominates in almost all of Europe. The early and complete adoption of the cremation rite in the Benelux area, as depicted in Radon-B, clearly sets it apart from the other regions. England, on the other hand, with its slow, gradual transition from inhumation to cremation, stands out due to its consistently moderate distances to all other regions. It remains without accentuated minima and maxima. The situation for *burial construction* is more complicated, as can be seen from the plots in the bottom right corner of Figure \ref{sed_region_matrix}. The distance does not seem to follow a general trend, but a multitude of local and temporally limited phenomena. This heterogeneity is of course also due to how correctly or incorrectly the proxy developed from Radon-B depicts developments in the real world. In the Middle Bronze Age, the Tumulus culture should bring about a fundamental reduction of the SED for almost all region-region relationships -- this effect is indeed weakly visible.

Table \ref{tab:distanceandcorrelationtable*} shows that the distance networks of *burial type* and *burial construction* are not equivalent. There is also no significant negative correlation. The degree of agreement between two regions in the question of inhumation and cremation is therefore not suitable for predicting the similarity of developments in these regions concerning the dominance of flat or mound graves. Only for the time slot at the transition to the Middle Bronze Age from 1800 to 1600calBC the 5% significance level of the Mantle test is almost undercut, thus suggesting a statistically significant relationship.

For the relationship between the SED network based on *burial type* and the spatial distance several observations result from Table \ref{tab:distanceandcorrelationtable*}: Cultural distance calculated from these to variables does not seem to correlate with spatial distance. A significant correlation between the distance measures only reveals itself for the Early Bronze Age time slot 2000 to 1800calBC: Only in the Early Bronze Age, long before the universal switch to the Urnfield burial tradition, large regions directly adjacent to each other tend to undergo the same developments rather than the ones further away from each other. In the Middle Bronze Age the SED developments of the *burial type* variable undergo the peaks already observed in Figure \ref{sed_region_matrix}. The correlation between cultural and spatial distance over these peaks is low and even tilts between 1400 and 1200calBC, at the transition to the Late Bronze Age, into the negative -- although not significantly. The cultural distance between several geographically distant regions is then smaller than between several directly adjacent regions. Looking at the SED network of *burial construction* (Table \ref{tab:distanceandcorrelationtable*}), similar conclusions can be drawn as in the *burial type* context: The distance values here are larger overall, and the SED differences within the individual spatial distance classes are clearly more accentuated. As well as in *burial type* the Early Bronze Age is by far the phase with the most substantial spatial correlation, though it is only significant for the time slot between 1800 and 1400calBC. There is no time period with negative mean correlation for *burial construction* cultural distance and spatial distance, but also here the correlation coefficient approaches zero in the Late Bronze Age.

The results of the simulation as presented in Figure \ref{simulation} confirm the expectation: In case of equal interaction intensity $I$ of separate populations, the correlation of cultural and spatial distance is weak (Figure \ref{simulation} A vs. C vs. E). It appears to be normally distributed around zero for many simulation runs, although individual runs can still yield remarkably high positive and negative correlation for certain time slots. The fluctuation of the correlation within one run can be strong and seems to increase for bigger values of $m_i$. If the interaction matrix $I$ is predefined to represent the spatial distance matrix (Figure \ref{simulation} BDF), then the mean correlation per time slot increases. The correlation increases as well with the degree of interaction $m_i$ (Figure \ref{simulation} D vs. F). Counter-intuitively the fluctuation of the correlation is also very high here (Figure \ref{simulation} E vs. F). The correlation development observed for the variables *burial type* and *burial construction* fits to simulation runs in all presets, but becomes less likely if we assume $I$ to represent spatial distance and a high degree of interaction $m_i$ (Figure \ref{simulation} A vs. F). That's mostly because of the relatively strong negative correlation of cultural and spatial distance in the Middle and Late Bronze Age (1400 to 1000calBC).

>>>>

In the course of the Bronze Age in Europe there was a trend away from body burial towards cremation. Around 2200 calBC cremation burials were almost unknown in Poland, Southern Germany, Northeastern France and Northern Germany. In the Northwest, in Great Britain and the Benelux area, and in the Southeast, in Austria and the Czech Republic, however, there were early cremation traditions. These contexts could be discussed as areas of origin of the later omnipresent phenomenon. While cremation burials lost importance in Austria and the Czech Republic as well as in the Benelux region during the Early Bronze Age and only regained relevance there in the Middle Bronze Age, their share increased steadily in England. Inhumation and cremation were balanced here for a long time. The situation was similar in Southern Scandinavia, where the proportion of cremation burials was much lower until the Late Bronze Age. In Poland and northern Germany, cremation became relatively suddenly and completely established with the beginning of the Late Bronze Age. In Northeastern France and Northern Germany this change took place already in the Middle Bronze Age. In both contexts inhumation continued to play an important role.

@@

Beyond the simple comparison of archaeological knowledge and information derived from Radon-B, Figure \ref{sed_region_matrix} shows the time series of squared Euclidean distance for each region-region relationship. The majority of the curves in its top right half for *burial type* correspond to expectations: At the beginning of the Bronze Age, inhumation is widespread throughout the study area, so the Euclidean distance at 2200 calBC is small in almost every region-region relationship. Only the Benelux countries, where the share of cremation is already high in the Early Bronze Age, are excluded. From the Middle Bronze Age onwards, but especially at the transition to the Late Bronze Age, the transition to the dominance of cremation took place all over Europe. The SED values increase in this period and show clear peaks or several successive maxima. This rise is a consequence of the time lag in the transition processes as depicted in Radon-B: It is different in every region. At the end of the Late Bronze Age, the average SED is approaching zero again, as burial by cremation dominates in almost all of Europe. The early and complete adoption of the cremation rite in the Benelux area, as depicted in Radon-B, clearly sets it apart from the other regions. England, on the other hand, with its slow, gradual transition from inhumation to cremation, stands out due to its consistently moderate distances to all other regions. It remains without accentuated minima and maxima. The situation for *burial construction* is more complicated, as can be seen from the plots in the bottom right corner of Figure \ref{sed_region_matrix}. The distance does not seem to follow a general trend, but a multitude of local and temporally limited phenomena. This heterogeneity is of course also due to how correctly or incorrectly the proxy developed from Radon-B depicts developments in the real world. In the Middle Bronze Age, the Tumulus culture should bring about a fundamental reduction of the SED for almost all region-region relationships -- this effect is indeed weakly visible.

Table \ref{tab:distanceandcorrelationtable*} shows that the distance networks of *burial type* and *burial construction* are not equivalent. There is also no significant negative correlation. The degree of agreement between two regions in the question of inhumation and cremation is therefore not suitable for predicting the similarity of developments in these regions concerning the dominance of flat or mound graves. Only for the time slot at the transition to the Middle Bronze Age from 1800 to 1600 calBC the 5% significance level of the Mantle test is almost undercut, thus suggesting a statistically significant relationship.

For the relationship between the SED network based on *burial type* and the spatial distance several observations result from Table \ref{tab:distanceandcorrelationtable*}: Cultural distance calculated from these to variables does not seem to correlate with spatial distance. A significant correlation between the distance measures only reveals itself for the Early Bronze Age time slot 2000 to 1800 calBC: Only in the Early Bronze Age, long before the universal switch to the Urnfield burial tradition, large regions directly adjacent to each other tend to undergo the same developments rather than the ones further away from each other. In the Middle Bronze Age the SED developments of the *burial type* variable undergo the peaks already observed in Figure \ref{sed_region_matrix}. The correlation between cultural and spatial distance over these peaks is low and even tilts between 1400 and 1200 calBC, at the transition to the Late Bronze Age, into the negative -- although not significantly. The cultural distance between several geographically distant regions is then smaller than between several directly adjacent regions. Looking at the SED network of *burial construction* (Table \ref{tab:distanceandcorrelationtable*}), similar conclusions can be drawn as in the *burial type* context: The distance values here are larger overall, and the SED differences within the individual spatial distance classes are clearly more accentuated. As well as in *burial type* the Early Bronze Age is by far the phase with the most substantial spatial correlation, though it is only significant for the time slot between 1800 and 1400 calBC. There is no time period with negative mean correlation for *burial construction* cultural distance and spatial distance, but also here the correlation coefficient approaches zero in the Late Bronze Age.

The results of the simulation as presented in Figure \ref{simulation} confirm the expectation: In case of equal interaction intensity $I$ of separate populations, the correlation of cultural and spatial distance is weak (Figure \ref{simulation} A vs. C vs. E). It appears to be normally distributed around zero for many simulation runs, although individual runs can still yield remarkably high positive and negative correlation for certain time slots. The fluctuation of the correlation within one run can be strong and seems to increase for bigger values of $m_i$. If the interaction matrix $I$ is predefined to represent the spatial distance matrix (Figure \ref{simulation} BDF), then the mean correlation per time slot increases. The correlation increases as well with the degree of interaction $m_i$ (Figure \ref{simulation} D vs. F). Counter-intuitively the fluctuation of the correlation is also very high here (Figure \ref{simulation} E vs. F). The correlation development observed for the variables *burial type* and *burial construction* fits to simulation runs in all presets, but becomes less likely if we assume $I$ to represent spatial distance and a high degree of interaction $m_i$ (Figure \ref{simulation} A vs. F). That's mostly because of the relatively strong negative correlation of cultural and spatial distance in the Middle and Late Bronze Age (1400 to 1000 calBC).

### 2019-02-02 article/04_Discussion.Rmd
more consistent spacing before "calBC" (@yesdavid)

`r a("Poland")`:
The region *Poland* completely covers the centre and western part of today's Poland. However, the ^14^C data from this region are almost exclusively from the voivodships of Kajuwien-Pommern, Lodsch and Lower Silesia. The amount of data is small, with significantly more data from the Early Bronze Age up to about 1700calBC. Almost no information is available from the Middle and Late Bronze Ages. This lack of dates applies to both the *burial type* and the *burial construction* variables, the former being slightly better documented. The relative development of *burial type* approximates the *real* conditions -- i.e. according to the classical, archaeological narrative: inhumations were significantly more frequent in western Poland until the Late Bronze Age, while cremations did rarely occur before the Urnfield Period. In contrast, it misrepresents *burial construction*: burial mounds also played an important role in Middle Bronze Age Poland [@czebreszukBronzeAgePolish2013; @dabrowskiAeltereBronzezeitPolen2004].

@@

`r a("Northern Germany")`:
The artificial region *Northern Germany* includes all federal states in East Germany, but also large parts of Lower Saxony and Schleswig-Holstein. The ^14^C data for this region come mainly from Saxony-Anhalt and neighbouring regions in Thuringia, Saxony and Lower Saxony. Also from the area north of Bremen some data are documented in Radon-B. The database does not contain as many dated graves from any other region as from Northern Germany. Most of them date back to the Early Bronze Age, while the Middle Bronze Age from about 1800calBC to 1400calBC seems barren. Only for the Late Bronze Age, there is a significant number of data documented again. The meta information on *burial type* and *burial construction* trace this development to the same extent. Since in Northern Germany influences from all four cardinal directions come together and contribute significantly to the formation of local cultural phenomena, it is difficult to estimate the overall development of the primary variables. The ^14^C proxy correctly represents *burial site* insofar as inhumations dominated in the Early Bronze Age and were later on replaced by cremations especially in the North from the Middle Bronze Age onwards. In the Late Bronze Age though, cremation was more dominant than the dataset suggests. Concerning *burial construction*, the current Radon-B stock records the fundamental dominance of flat graves well, but attributes more relevance to them in the Early and Middle Bronze Age than it is usual in the archaeological assessment of the context [@hofmannRituelleUmgangMit2008; @jockenhovelGermanyBronzeAge2013].

`r a("Southern Scandinavia")`:
The region *Southern Scandinavia* includes all of Denmark with Jutland, Funen, Zeeland and all smaller islands as well as large parts of East of Southern Sweden, and coastal regions in Schleswig-Holstein and Mecklenburg-Western Pomerania. The ^14^C data are very well distributed across Jutland, Funen and the Swedish Skåne region. Since the Bronze Age in Scandinavia begins later than in Central Europe and is measured by another chronological system, the diachronic developments cannot merely be compared through the tripartite structure of the Early, Middle and Late Bronze Ages. The period which is labelled Early Bronze Age in Central Europe corresponds to the Scandinavian Late Neolithic. The majority of data from Scandinavia are from the period IB onwards. In this respect, the data density for almost the entire Bronze Age is good -- also with regard to the two variables *burial type* and *burial construction*. The Late Neolithic from 2200calBC is a desideratum. If one carries out the necessary mental shift -- the Central European Middle Bronze Age corresponds to the Nordic Early Bronze Age -- and compares the proxy of the relative grave form development with the archaeological observations, then one finds a substantial agreement: inhumations were the rule in periods I and II (about 1800-1300calBC), but cremations also occurred sporadically. From period III onwards cremation burials gained the upper hand. In the Scandinavian Early Bronze Age burials in burial mounds dominated. In the Late Bronze Age, flat grave fields began to occur [@thraneScandinavia2013].

@@

Against this background, the observation that there is no permanent, significant correlation either between the variables or with the spatial distance is all the more astonishing (Table \ref{tab:distanceandcorrelationtable*}). Only for small periods of time this correlation can be assumed based on the Radon-B data. The *burial construction* distance shows a higher degree of correlation with spatial distance than the *burial type* distance, which even seems to behave in an opposing manner: negative correlation. If one assumes the reconstructed timelines of cultural similarity intensity to be correct, than one has to explain a rise (between 2200 and 2000calBC) and a decline (after 2000) of cultural-spatial-distance correlation. There are several reasons why any interpretation of this observation is difficult so far:

>>>>

`r a("Poland")`:
The region *Poland* completely covers the centre and western part of today's Poland. However, the ^14^C data from this region are almost exclusively from the voivodships of Kajuwien-Pommern, Lodsch and Lower Silesia. The amount of data is small, with significantly more data from the Early Bronze Age up to about 1700 calBC. Almost no information is available from the Middle and Late Bronze Ages. This lack of dates applies to both the *burial type* and the *burial construction* variables, the former being slightly better documented. The relative development of *burial type* approximates the *real* conditions -- i.e. according to the classical, archaeological narrative: inhumations were significantly more frequent in western Poland until the Late Bronze Age, while cremations did rarely occur before the Urnfield Period. In contrast, it misrepresents *burial construction*: burial mounds also played an important role in Middle Bronze Age Poland [@czebreszukBronzeAgePolish2013; @dabrowskiAeltereBronzezeitPolen2004].

@@

`r a("Northern Germany")`:
The artificial region *Northern Germany* includes all federal states in East Germany, but also large parts of Lower Saxony and Schleswig-Holstein. The ^14^C data for this region come mainly from Saxony-Anhalt and neighbouring regions in Thuringia, Saxony and Lower Saxony. Also from the area north of Bremen some data are documented in Radon-B. The database does not contain as many dated graves from any other region as from Northern Germany. Most of them date back to the Early Bronze Age, while the Middle Bronze Age from about 1800 to 1400 calBC seems barren. Only for the Late Bronze Age, there is a significant number of data documented again. The meta information on *burial type* and *burial construction* trace this development to the same extent. Since in Northern Germany influences from all four cardinal directions come together and contribute significantly to the formation of local cultural phenomena, it is difficult to estimate the overall development of the primary variables. The ^14^C proxy correctly represents *burial site* insofar as inhumations dominated in the Early Bronze Age and were later on replaced by cremations especially in the North from the Middle Bronze Age onwards. In the Late Bronze Age though, cremation was more dominant than the dataset suggests. Concerning *burial construction*, the current Radon-B stock records the fundamental dominance of flat graves well, but attributes more relevance to them in the Early and Middle Bronze Age than it is usual in the archaeological assessment of the context [@hofmannRituelleUmgangMit2008; @jockenhovelGermanyBronzeAge2013].

`r a("Southern Scandinavia")`:
The region *Southern Scandinavia* includes all of Denmark with Jutland, Funen, Zeeland and all smaller islands as well as large parts of East of Southern Sweden, and coastal regions in Schleswig-Holstein and Mecklenburg-Western Pomerania. The ^14^C data are very well distributed across Jutland, Funen and the Swedish Skåne region. Since the Bronze Age in Scandinavia begins later than in Central Europe and is measured by another chronological system, the diachronic developments cannot merely be compared through the tripartite structure of the Early, Middle and Late Bronze Ages. The period which is labelled Early Bronze Age in Central Europe corresponds to the Scandinavian Late Neolithic. The majority of data from Scandinavia are from the period IB onwards. In this respect, the data density for almost the entire Bronze Age is good -- also with regard to the two variables *burial type* and *burial construction*. The Late Neolithic from 2200 calBC is a desideratum. If one carries out the necessary mental shift -- the Central European Middle Bronze Age corresponds to the Nordic Early Bronze Age -- and compares the proxy of the relative grave form development with the archaeological observations, then one finds a substantial agreement: inhumations were the rule in periods I and II (about 1800-1300 calBC), but cremations also occurred sporadically. From period III onwards cremation burials gained the upper hand. In the Scandinavian Early Bronze Age burials in burial mounds dominated. In the Late Bronze Age, flat grave fields began to occur [@thraneScandinavia2013].

@@

Against this background, the observation that there is no permanent, significant correlation either between the variables or with the spatial distance is all the more astonishing (Table \ref{tab:distanceandcorrelationtable*}). Only for small periods of time this correlation can be assumed based on the Radon-B data. The *burial construction* distance shows a higher degree of correlation with spatial distance than the *burial type* distance, which even seems to behave in an opposing manner: negative correlation. If one assumes the reconstructed timelines of cultural similarity intensity to be correct, than one has to explain a rise (between 2200 and 2000 calBC) and a decline (after 2000 calBC) of cultural-spatial-distance correlation. There are several reasons why any interpretation of this observation is difficult so far:
